---
title: "Functional Analysis"
author: "Sergey Naumenko"
date: "`r Sys.Date()`"
output:
   html_document:
      code_folding: hide
      df_print: paged
      highlights: pygments
      number_sections: true
      self_contained: true
      theme: default
      toc: true
      toc_float:
         collapsed: true
         smooth_scroll: true
---

# Overview

- Principal Investigator: Konstantina Stankovic
- Researcher: Minjin Jeong
- Experiment: RNAseq and drug repurposing analysis of mutant TMPRSS3 iPSC line [hbc03750]
- more about functional analysis: https://hbctraining.github.io/DGE_workshop_salmon/lessons/functional_analysis_2019.html

```{r setup, message = FALSE}
library(tidyverse)
library(clusterProfiler)
library(topGO)
library(DOSE)
library(pathview)
library(org.Hs.eg.db)
library(AnnotationHub)
library(GSEABase)
library(readxl)
library(writexl)

# functions
# genes2test - up or down ensembl_gene_id vector
# all_genes - all genes ensembl_gene_id_vector
# ont = ["BP", "CC", "MF"]
# returns: enrichResult
go_analysis <- function(genes2test, all_genes, ont){
    ego <- enrichGO(
            gene = genes2test,
            universe = all_genes,
            OrgDb = org.Hs.eg.db,
            keyType = "ENSEMBL",
            ont = ont,
            qvalueCutoff = 0.05,
            readable = TRUE) 
    return(ego)
}

gsea_analysis <- function(genes2test, all_genes){
    genes <- genes_df$log2FoldChange
    names(genes) <- genes_df$entrezgene_id
    genes <- sort(genes, decreasing = T)

    ewp2 <- GSEA(genes, 
             TERM2GENE = wpid2gene, 
             TERM2NAME = wpid2name, 
             verbose = FALSE)

    ewp2 <- setReadable(ewp2, org.Hs.eg.db, keyType = "ENTREZID")
    ewp2 <- ewp2 %>% as.data.frame()
    return(ewp2)
}

#hub <- AnnotationHub()
# query(hub, "Homo sapiens")
#hsapiens <- hub[["AH75011"]]
```    

# Wiki Pathways analysis
Pathway analysis is for up- and down- regulated genes together as genes could work in a pathway both ways.
```{r wiki_pathways, rows.print = 20}
# input: entrez_id vector
wiki_pathway_analysis <- function(genes){
    wp2gene <- read.gmt("tables/wikipathways-20191110-gmt-Homo_sapiens.gmt")
    wp2gene <- wp2gene %>% separate(ont, c("name", "version", "wpid", "org"), "%")
    wpid2gene <- wp2gene %>% dplyr::select(wpid, gene) #TERM2GENE
    wpid2name <- wp2gene %>% dplyr::select(wpid, name) #TERM2NAME
    ewp <- enricher(genes, 
                TERM2GENE = wpid2gene, 
                TERM2NAME = wpid2name)
    ewp <- setReadable(ewp, org.Hs.eg.db, keyType = "ENTREZID")
    ewp_df <- ewp %>% as.data.frame()
    return(ewp_df)
}

genes_up <- read_xlsx("tables/genes_up.xlsx") %>% 
        drop_na(entrezgene_id) %>% 
        dplyr::filter(log2FoldChange >= 1)

ewp_df <- wiki_pathway_analysis(genes_up$entrezgene_id)

ewp_df
```

```{r}
genes_down <- read_xlsx("tables/genes_down.xlsx") %>% 
        drop_na(entrezgene_id) %>% 
        dplyr::filter(log2FoldChange <= -1)

ewp_df <- wiki_pathway_analysis(genes_down$entrezgene_id)
ewp_df
```

# Gene ontology enrichment analysis
## Upregulated genes
```{r go_enrichment_bp, fig.width = 15, fig.height = 15, include = FALSE}
all_genes <- read_csv("tables/genes_all.csv")
go_result <- go_analysis(genes_up$ensembl_gene_id, all_genes$ensembl_gene_id, "BP")
df_go_up_bp <- as.data.frame(go_result)
```

### Biological process
```{r, fig.width = 10, fig.height = 10}
dotplot(go_result, showCategory = 50) + ggtitle ("Biological process")
```
```{r fig.width = 10, fig.height = 10}
emapplot(go_result, showCategory = 50)
```
```{r, fig.width = 10, fig.height = 10}
foldchanges <- genes_up$log2FoldChange
names(foldchanges) <- genes_up$external_gene_name

cnetplot(go_result,
         categorySize="pvalue", 
         showCategory = 5, 
         foldChange = foldchanges, 
         vertex.label.font = 6)
```

### Cellular compartment

```{r go_enrichment_cc, fig.width = 10, fig.height = 10}
go_result <- go_analysis(genes_up$ensembl_gene_id, all_genes$ensembl_gene_id, "CC")
df_go_up_cc <- as.data.frame(go_result)
dotplot(go_result, showCategory = 50) + ggtitle ("Cellular compartment")
```

```{r fig.width = 10, fig.height = 10}
emapplot(go_result, showCategory = 50)
```

```{r, fig.width = 10, fig.height = 10}
cnetplot(go_result,
         categorySize = "pvalue", 
         showCategory = 5, 
         foldChange = foldchanges, 
         vertex.label.font = 6)
```

### Molecular function

```{r go_enrichment_mf, fig.width = 10, fig.height = 10}
go_result <- go_analysis(genes_up$ensembl_gene_id, all_genes$ensembl_gene_id, "MF")
df_go_up_mf <- as.data.frame(go_result)
dotplot(go_result, showCategory = 50) + ggtitle ("Molecular function")
```

```{r fig.width = 10, fig.height = 10}
emapplot(go_result, showCategory = 50)
```

```{r, fig.width = 10, fig.height = 10}
cnetplot(go_result,
         categorySize = "pvalue", 
         showCategory = 5, 
         foldChange = foldchanges, 
         vertex.label.font = 6)
```

## Downregulated genes - no enrichment

### Biological process
```{r go_enrichment_down_bp, fig.width = 10, fig.height = 10}
go_result <- go_analysis(genes_down$ensembl_gene_id, 
                         all_genes$ensembl_gene_id, 
                         "BP")
df_go_down_bp <- as.data.frame(go_result)
if (nrow(df_go_down_bp) > 0){
    dotplot(go_result, showCategory = 50) + ggtitle ("Biological process")
}
```

```{r fig.width = 10, fig.height = 10}
if (nrow(df_go_down_bp) > 0){
    emapplot(go_result, showCategory = 50)
}
```

```{r, fig.width = 10, fig.height = 10}
if (nrow(df_go_down_bp) > 0){
    foldchanges <- genes_down$log2FoldChange
    names(foldchanges) <- genes_down$external_gene_name

    cnetplot(go_result,
         categorySize = "pvalue", 
         showCategory = 5, 
         foldChange = foldchanges, 
         vertex.label.font = 6)
}
```

### Cellular compartment

```{r go_enrichment_down_cc, fig.width = 10, fig.height = 10}
go_result <- go_analysis(genes_down$ensembl_gene_id, 
                         all_genes$ensembl_gene_id, 
                         "CC")
df_go_down_cc <- as.data.frame(go_result)
if (nrow(df_go_down_cc) > 0){
    dotplot(go_result, showCategory = 50) + ggtitle ("Cellulcar compartment")
}
```

```{r fig.width = 10, fig.height = 10}
if (nrow(df_go_down_cc) > 0){
    emapplot(go_result, showCategory = 50)
}
```

```{r, fig.width = 10, fig.height = 10}
if (nrow(df_go_down_cc) > 0){
    foldchanges <- genes_down$log2FoldChange
    names(foldchanges) <- genes_down$external_gene_name

    cnetplot(go_result,
         categorySize="pvalue", 
         showCategory = 5, 
         foldChange = foldchanges, 
         vertex.label.font = 6)
}
```

### Molecular function

```{r go_enrichment_down_mf, fig.width = 10, fig.height = 10}
go_result <- go_analysis(genes_down$ensembl_gene_id, all_genes$ensembl_gene_id, "MF")
df_go_down_mf <- as.data.frame(go_result)
if (nrow(df_go_down_mf) > 0){
    dotplot(go_result, showCategory = 50) + ggtitle ("Molecular function")
}
```

```{r fig.width = 10, fig.height = 10}
if (nrow(df_go_down_mf) > 0){
    emapplot(go_result, showCategory = 50)
}
```

```{r, fig.width = 10, fig.height = 10}
if (nrow(df_go_down_mf) > 0){
    foldchanges <- genes_down$log2FoldChange
    names(foldchanges) <- genes_down$external_gene_name

    cnetplot(go_result,
         categorySize="pvalue", 
         showCategory = 5, 
         foldChange = foldchanges, 
         vertex.label.font = 6)
}
```

```{r}
write_xlsx(list(df_go_up_bp, df_go_up_cc, df_go_up_mf,
                df_go_down_bp, df_go_down_cc, df_go_down_mf), 
       "tables/go_enrichment.xlsx")
```

# Gene set enrichment analysis
```{r}
## Remove any NA values (reduces the data by quite a bit)
res_entrez <- all_genes %>% 
            drop_na(entrezgene_id) %>% 
            distinct(entrezgene_id, .keep_all = TRUE)

foldchanges <- res_entrez$log2FoldChange
names(foldchanges) <- res_entrez$entrezgene_id
foldchanges <- sort(foldchanges, decreasing = TRUE)

gseaKEGG <- gseKEGG(geneList = foldchanges, 
                    organism = "hsa", 
                    nPerm = 1000, 
                    minGSSize = 10,
                    verbose = FALSE,
                    )

gseaKEGG <- setReadable(gseaKEGG, org.Hs.eg.db, keyType = "ENTREZID")

## Extract the GSEA results
gseaKEGG_results <- gseaKEGG@result

write_xlsx(gseaKEGG_results, "tables/gsea_kegg.xlsx")
View(gseaKEGG_results)
```

## Top enriched set

```{r}
gseaplot(gseaKEGG, "hsa04080")
```

```{r}
detach("package:dplyr", unload=TRUE) # first unload dplyr to avoid conflicts

## Output images for a single significant KEGG pathway
pathview(gene.data = foldchanges,
              pathway.id = "hsa04080",
              species = "hsa",
              limit = list(gene = 2, # value gives the max/min limit for foldchanges
              cpd = 1))
```

```{r}
all_genes_gsea <- all_genes %>% dplyr::select(external_gene_name, ensembl_gene_id, 
                                       SK7A2, SK7A3, SK7A4, 
                                       SK8A2, SK8A3, SK8A4) %>% 
                            tidyr::drop_na(external_gene_name) %>% 
                            dplyr::distinct(external_gene_name, .keep_all = TRUE) %>% 
                            dplyr::rename(NAME = external_gene_name,
                                          DESCRIPTION = ensembl_gene_id)
write_tsv(all_genes_gsea, "tables/all_genes_gsea.txt")
                            
#ewp_df <- gsea_analysis(genes_up)
#ewp_df <- gsea_analysis(genes_down)
```

# R Session
```{r}
sessionInfo()
```
