---
title: "Marker identification"
author: "Sergey Naumenko"
date: "`r Sys.Date()`"
output:
    html_document:
        code_folding: hide
        df_print: paged
        highlight: tango
        theme: default
        number_sections: true
        toc: true
        toc_float:
            collapsed: true
            smooth_scroll: false
params:
    variable_features: 3000
    npcs: 30
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(readxl)
library(writexl)
library(Seurat)

ggplot2::theme_set(theme_light(base_size = 14))

opts_chunk[["set"]](
    cache = FALSE,
    dev = c("png", "pdf"),
    error = TRUE,
    highlight = TRUE,
    message = FALSE,
    prompt = FALSE,
    tidy = FALSE,
    warning = FALSE)

```

# Overview
- Principal Investigator: 
- Researcher: 
- Experiment: 
- [I used this lesson as a template](https://github.com/hbctraining/scRNA-seq/blob/master/lessons/07_SC_marker_identification.md)
- markers annotated with [brainrnaseq](https://www.brainrnaseq.org/)

# All markers

Description of the columns:

- cluster: cluster ID
- gene: ENSEMBL gene id
- external_gene_name: gene name
- description: gene_description
- p_val_adj: Bonferroni-corrected p-value
- p_val: raw p-value
- avg_logFC: average log2 fold change. Positive values indicate that the gene is more highly expressed in the cluster.
- pct.1: The percentage of cells where the gene is detected in the cluster
- pct.2: The percentage of cells where the gene is detected on average in the other clusters

```{r, rows.print = 25}
seurat <- readRDS("data/seurat.clusters.RDS")
markers_file <- "tables/markers.annotated.xlsx"
if (!file.exists("markers_file")){
    # long (~1h)
    markers <- FindAllMarkers(object = seurat,
                          only.pos = TRUE,
                          logfc.threshold = 0.25,
                          min.pct = 0.25
                          )
    ensembl_w_description <- read_csv("tables/ensembl_w_description.mouse.protein_coding.csv")

    #combine markers with gene descriptions
    ann_markers <- inner_join(x = markers,
                          y = ensembl_w_description,
                          by = c("gene" = "ensembl_gene_id")) %>% unique()

    # Order the rows by p-adjusted values
    ann_markers <- ann_markers %>%
            dplyr::arrange(cluster, p_val_adj) %>% 
        select(cluster, gene, external_gene_name, description, p_val_adj, p_val, avg_logFC, pct.1, pct.2)

    write_xlsx(list(ann_markers), markers_file)
}else{
    ann_marders <- read_xlsx(markers_file)
}
ann_markers

```
[Open the annontated markers table in Excel](`r markers_file`)

# R session information
```{r}
sessionInfo()
```
