---
title: "Clustering"
author: "Sergey Naumenko"
date: "`r Sys.Date()`"
output:
    html_document:
        code_folding: hide
        df_print: paged
        highlight: tango
        theme: default
        number_sections: true
        toc: true
        toc_float:
            collapsed: true
            smooth_scroll: false
params:
    variable_features: 2000
    heavy_calc: FALSE
---

# Overview

- Researcher: 
- Experiment: 
- variable_features: `r params$variable_features`

```{r setup, include = FALSE}
library(knitr)
library(cowplot)
library(tidyverse)
library(Seurat)
library(RCurl)
library(Rtsne)

ggplot2::theme_set(theme_light(base_size = 14))

opts_chunk[["set"]](
    cache = FALSE,
    dev = c("png", "pdf"),
    error = TRUE,
    highlight = TRUE,
    message = FALSE,
    prompt = FALSE,
    tidy = FALSE,
    warning = FALSE)

```

```{r}
pbmc_data <- Read10X(data.dir = "data/pbmc3k_filtered_gene_bc_matrices/filtered_gene_bc_matrices/hg19/")
pbmc <- CreateSeuratObject(counts = pbmc_data, 
                           project = "pbmc3k", 
                           min.cells = 3, 
                           min.features = 200)

pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")

VlnPlot(pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

```{r}
plot1 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
CombinePlots(list(plot1, plot2))
```

```{r}
pbmc <- subset(pbmc, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)
```


# Normalization
```{r normalization}
# Normalize, Scale, Find Variable Features
pbmc <- SCTransform(pbmc, vars.to.regress = "percent.mt")
```

# Clustering
```{r}
pbmc <- RunPCA(pbmc, verbose = FALSE)
pbmc <- RunUMAP(pbmc, dims = 1:30, verbose = FALSE)
pbmc <- FindNeighbors(pbmc, dims = 1:30, verbose = FALSE)
pbmc <- FindClusters(pbmc, verbose = FALSE)
DimPlot(pbmc, label = TRUE) + NoLegend()
```


# Top variable genes
```{r, fig.width = 20, fig.height = 10}
pbmc <- FindVariableFeatures(pbmc, selection.method = "vst", nfeatures = 2500)
top20 <- head(VariableFeatures(pbmc), 20)
plot1 <- VariableFeaturePlot(pbmc)
plot2 <- LabelPoints(plot = plot1, point = top20, repel = TRUE)
CombinePlots(plots = list(plot1, plot2))
```

# SCTransform
```{r scaling}
all.genes <- rownames(pbmc)
pbmc <- ScaleData(pbmc, features = all.genes)
# or to scale only VariableFeatures:
# pbmc <- ScaleData(pbmc)
```

[Open cell cycle genes in excel](tables/musculus_cell_cycle.csv)

```{r}
s_genes <- cell_cycle_genes %>% 
    dplyr::filter(phase == "S") %>% 
    pull(geneID)

g2m_genes <- cell_cycle_genes %>% 
    dplyr::filter(phase == "G2/M") %>% 
    pull(geneID)

if (params$heavy_calc){
  seurat <- CellCycleScoring(seurat,
                         g2m.features = g2m_genes,
                         s.features = s_genes)
  # heavy, memory intense
  seurat <- RunPCA(seurat)
  saveRDS(seurat, paste0(filter_path, "seurat.pca1.RDS"))
}else{
  seurat <- readRDS(paste0(filter_path, "seurat.pca1.RDS"))
}

DimPlot(seurat,
        reduction = "pca",
        group.by = "Phase") +
    ggtitle("PCA before regression of covariates")

```

```{r pca2}
# Define variables in metadata to regress
vars_to_regress <- c("nUMI", "S.Score", "G2M.Score")

# Regress out the uninteresting sources of variation in the data
# heavy
if (params$heavy_calc){
    seurat <- ScaleData(object = seurat,
                        vars.to.regress = vars_to_regress, 
                        verbose = FALSE)

    # Re-run the PCA - heavy
    seurat <- RunPCA(object = seurat)
    saveRDS(seurat, paste0(filter_path, "seurat.pca2.RDS")) 
}else{
    seurat <- readRDS(paste0(filter_path, "seurat.pca2.RDS"))
}

DimPlot(object = seurat,
        reduction = "pca",
        group.by = "Phase") + 
    ggtitle("PCA after regression")

```

# Usable principal components for t-SNE and UMAP

## Identify significant PCs
```{r}
# ugly plot, needs fixing
#DimHeatmap(seurat,
#    dims = 2,
#    cells = 100,
#    balanced = TRUE) +
#    ggtitle("Dimension 2")
```

```{r}
print(x = seurat[["pca"]],
      dims = 1:10,
      nfeatures = 5)
```

## Elbow plot: how many PCs to use
```{r}
ElbowPlot(object = seurat,
          ndims = 30) + 
    geom_vline(xintercept = 13.5)
```

```{r}
pct <- seurat[["pca"]]@stdev / sum(seurat[["pca"]]@stdev) * 100
# Calculate cumulative percents for each PC
cumu <- cumsum(pct)
# Determine which PC exhibits cumulative percent greater than 90% and % variation associated with the PC as less than 5
co1 <- which(cumu > 90 & pct < 5)[1]
co2 <- sort(which((pct[1:length(pct) - 1] - pct[2:length(pct)]) > 0.1), decreasing = T)[1] + 1
# last point where change of % of variation is more than 0.1%.
pc_threshold <- min(co1, co2)
```

PC threshold: `r pc_threshold`

```{r}
plot_df <- data.frame(pct = pct, cumu = cumu, rank = 1:length(pct))

ggplot(plot_df, aes(cumu, pct, label = rank, color = rank > pc_threshold)) + 
    geom_text() + 
    geom_vline(xintercept = 90, color = "grey") + 
    geom_hline(yintercept = min(pct[pct > 5]), color = "grey") + theme_bw() +
    ggtitle("Significant PCs")
```


# Cluster cells with t-SNE and UMAP
```{r findclusters, message }
if (params$heavy_calc){
    seurat <- FindNeighbors(object = seurat,
                          dims = 1:pc_threshold)

    # heavy - 40s
    seurat <- FindClusters(object = seurat,
                       resolution = c(0.4, 0.6, 0.8, 1.0, 1.2, 1.8))

    # small number of duplicates is ok    
    seurat <- RunTSNE(object = seurat,
                  check_duplicates = F)

    #for UMAP it is necessary to install python umap-learn
    seurat <- RunUMAP(seurat, 
                  reduction = "pca", 
                  dims = 1:pc_threshold)
  
    saveRDS(seurat, paste0(filter_path, "seurat.tsne_umap_res.RDS"))

}else{
    seurat <- readRDS(paste0(filter_path, "seurat.tsne_umap_res.RDS"))
}
```

```{r}
## t-SNE - resolution 0.4
#Idents(object = seurat) <- "RNA_snn_res.0.4"
#DimPlot(object = seurat, 
#        label = T,
#        reduction = "tsne",
#        plot.title = "t-SNE")
```


## UMAP - resolution 0.4
```{r}
Idents(object = seurat) <- "RNA_snn_res.0.4"

# Plot the UMAP
DimPlot(seurat,
        reduction = "umap",
        label = TRUE,
        label.size = 6,
        plot.title = "UMAP_RNA_snn_res.0.4")

```

## N cells in clusters - resolution 0.4
```{r, fig.width = 10, fig.height = 10}
n_cells <- FetchData(seurat,
                     vars = c("ident")) %>% 
            dplyr::count(ident) 

n_cells %>% ggplot(aes(x = ident, y = n)) + geom_col() +
    ggtitle("Number of cells in clusters")+
    geom_text(aes(label = n), hjust = 0) +
    coord_flip() +
    xlab("Cluster ID") +
    ylab("Number of cells")
```

# Clusters and cell cycle
```{r}
group_by <- c("Phase")

# Getting coordinates for cells to use for UMAP and associated grouping variable information
class_umap_data <- FetchData(seurat, 
                             vars = c("ident", "UMAP_1", "UMAP_2", group_by))

# Adding cluster label to center of cluster on UMAP
umap_label <- FetchData(seurat, 
                        vars = c("ident", "UMAP_1", "UMAP_2"))  %>%
        group_by(ident) %>%
        summarise(x = mean(UMAP_1), y = mean(UMAP_2))


# Getting coordinates for cells to use for PCA and associated grouping variable information
class_pca_data <- FetchData(seurat, 
                            vars = c("ident", "PC_1", "PC_2", group_by))

# Adding cluster label to center of cluster on PCA
pca_label <- FetchData(seurat, 
                       vars = c("ident", "PC_1", "PC_2"))  %>%
        mutate(ident = seurat@active.ident) %>%
        group_by(ident) %>%
        summarise(x = mean(PC_1), y = mean(PC_2))

```


```{r}
# Function to plot UMAP and PCA as grids
plot_grid(
      ggplot(class_umap_data, aes(UMAP_1, UMAP_2)) +
        geom_point(aes_string(color = group_by), alpha = 0.7) +
        scale_color_brewer(palette = "Set2")  +
        geom_text(data=umap_label, aes(label=ident, x, y)),
      ggplot(class_pca_data, aes(PC_1, PC_2)) +
        geom_point(aes_string(color = group_by), alpha = 0.7) +
        scale_color_brewer(palette = "Set2")  +
        geom_text(data=pca_label, 
                  aes(label=ident, x, y)),
      nrow = 1, 
      align = "v")
```

```{r}
# Determine metrics to plot present in seurat_control@meta.data
# no nUMI nGene mitoRatio
metrics <-  c("S.Score", "G2M.Score")

# Extract the UMAP coordinates for each cell and include information about the metrics to plot
qc_data <- FetchData(seurat, 
                     vars = c(metrics, "ident", "UMAP_1", "UMAP_2"))

# Plot a UMAP plot for each metric
map(metrics, function(qc){
        ggplot(qc_data,
               aes(UMAP_1, UMAP_2)) +
                geom_point(aes_string(color=qc), 
                           alpha = 0.7) +
                scale_color_gradient(guide = FALSE, 
                                     low = "grey90", 
                                     high = "blue")  +
                geom_text(data=umap_label, 
                          aes(label = ident, x, y)) +
                ggtitle(qc)
}) %>% plot_grid(plotlist = .)

```

# UMAP plot for each of the PCs
```{r}
# Defining the information in the seurat object of interest
columns <- c(paste0("PC_", 1:pc_threshold),
            "ident",
            "UMAP_1", "UMAP_2")

# Extracting this data from the seurat object
pc_data <- FetchData(seurat, 
                     vars = columns)

map(paste0("PC_", 1:pc_threshold), function(pc){
        ggplot(pc_data, 
               aes(UMAP_1, UMAP_2)) +
                geom_point(aes_string(color=pc), 
                           alpha = 0.7) +
                scale_color_gradient(guide = FALSE, 
                                     low = "grey90", 
                                     high = "blue")  +
                geom_text(data=umap_label, 
                          aes(label=ident, x, y)) +
                ggtitle(pc)
}) %>% plot_grid(plotlist = .)
```

# UMAP clusters colored by condition
```{r}
Idents(object = seurat) <- "RNA_snn_res.0.4"
DimPlot(seurat,
        reduction = "umap",
        group.by = "condition"
                ) +
    ggtitle("Clusters by condition")
```

# UMAP clusters colored by sample
```{r}
Idents(object = seurat) <- "RNA_snn_res.0.4"
DimPlot(seurat,
        reduction = "umap",
        group.by = "sample_id"
        ) +
    ggtitle("Clusters by sample")
```

## UMAP clusters for each sample
```{r, fig.width = 20, fig.height = 20}
Idents(object = seurat) <- "RNA_snn_res.0.4"

# Plot the UMAP
DimPlot(seurat,
        reduction = "umap",
        label = TRUE,
        label.size = 6,
        plot.title = "UMAP_RNA_snn_res.0.4",
        split.by = "sample_id",
        ncol = 4, 
        group.by = "condition")

```

# Conclusions:

# R session information
```{r}
sessionInfo()
```
